---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# AI-assisted module development {#sec-aiModules}

::: {.callout-tip appearance="simple"}
See the [Glossary](glossary.qmd) for definitions of recurring terms
(e.g. *SpaDES module*, *SpaDES event*, *interoperable model*,
*modular workflow*).
:::

## AI as an accelerator for a reusable module ecosystem {#sec-aiAccelerator}

The hardest part of growing a SpaDES ecosystem isn't writing any one
module -- it's keeping the *interfaces* consistent across many of them
so they actually compose. A module that calls a covariate
`pixelGroupMap` plugs into the same pipeline as `Biomass_core`; one
that calls it `pgMap` does not, even if the data are identical. The
metadata (`expectsInput`, `createsOutput`, parameter names, `timeunit`)
is what turns a clever script into a piece of shared infrastructure.

That metadata is largely mechanical to write -- and exactly the kind
of work an AI assistant handles well. Given a linear R script and a
few existing modules to mimic, a current LLM can translate the script
into a `defineModule` + `doEvent` + scheduling skeleton in seconds,
matched to the conventions used by `LandR`, `scfm`, `RSFpredict`, and
the rest. The role of the scientist shifts from *typing the
boilerplate* to *checking that the boilerplate is right* -- which
inputs are needed, which outputs are produced, which parameters belong
to which event.

There is a network effect: the more existing modules conform to a
shared pattern, the better an LLM can produce a *new* module that fits.
Today's `Biomass_core` becomes the template for tomorrow's
`Biomass_disturbance`; today's `RSFpredict` becomes the template for
the next species' SDM-as-module. The ecosystem builds itself faster,
and stays consistent because nobody has to *remember* the conventions
-- the AI re-applies them on every conversion.

This chapter is a worked example. We will take the linear R script we
built in [Forecasting wildlife habitat @sec-rsfForecast] and walk
through the prompts that turn it into the `RSFpredict` module. The
goal isn't to show off the prompts -- it's to show that the
conversion is now a *short conversation*, not a refactor project.

## Worked example: turning an RSF forecast into a SpaDES module

We start from the linear script in [Converting habitat model
prediction code into a SpaDES module @sec-rsfPredictModule]: one
setup section, one annual loop, no `sim$`, no `doEvent`. Four prompts
get us to a working module: one to scaffold the module's directory,
then three to fill in metadata, init, and the recurring events.

### Prompt 0 -- scaffold the module

Before any AI work, generate the standard module directory layout so
the AI has a *file* to edit rather than a clean slate to author from
scratch. `SpaDES.core::newModule()` writes the template:

```{r newModule0}
SpaDES.core::newModule(
  name = "RSFpredict",
  path = "~/SpaDES_book/RSF_forecast/modules",
  open = FALSE
)
```

The skeleton has `# ! ----- EDIT BELOW ----- !` markers showing where
each section goes -- the AI uses these as anchors and the diffs become
trivially reviewable. You could ask the AI to generate the whole file
in one shot instead, but starting from `newModule()`'s scaffold gives
you a stable target that already conforms to SpaDES conventions, and
keeps each prompt narrow.

### Prompt 1 -- fill in the metadata

> Here is a linear R script that forecasts a Resource Selection
> Function across a simulated landscape, year by year. I already ran
> `SpaDES.core::newModule("RSFpredict", path = ...)` so the scaffold
> exists at `modules/RSFpredict/RSFpredict.R`. Replace the
> placeholder `defineModule()` block in that file with a real one.
> Include:
>
> - the required packages (`reproducible`, `terra`, `glmmTMB`, plus the
>   ones the script uses);
> - `parameters` for every value I had in the `params` list (numBins,
>   simulationProcess, predictionInterval, predictStartYear,
>   predictLastYear, ts_else, .studyAreaName) -- declare type and a
>   sensible default for each;
> - `inputObjects` for the variables the script assumed were in
>   scope: `model` (`glmmTMB`), `studyArea` (`SpatVector`),
>   `modelLand`, `cohortData`, `pixelGroupMap`, `timeSinceFire`,
>   `rasterToMatch` (all `SpatRaster`);
> - `outputObjects` for `pred`, `binMap`, `simLand`, `simPred`,
>   `simBinMap`.
>
> Use `timeunit = "year"`. No event code yet -- just the metadata.

The AI's output is essentially the `defineModule(...)` block we showed
at [@sec-rsfPredictModule § Step 1] -- a declaration of what the
module needs, what it produces, and what its parameters are. The
**single most important thing to check** is that the `inputObjects`
names match what the upstream modules (`Biomass_core`, `scfm`)
*produce*. If they don't, the project won't wire up; if they do, you
get composition for free.

### Prompt 2 -- fill in the `init` event + baseline computation

> Now edit the `doEvent.RSFpredict()` stub `newModule()` already wrote.
> Keep the `switch()` shell but replace the `init` branch contents.
> The `init` event should:
>
> - schedule a `buildBaselineRSFmap` event at `time(sim)` (runs once,
>   immediately);
> - if `P(sim)$simulationProcess == "dynamic"`, also schedule
>   `simLayers` and `simRSFmap` events at `P(sim)$predictStartYear`;
> - if `P(sim)$predictLastYear` is `TRUE`, schedule a final
>   `simLayers` at `end(sim)`.
>
> Then write the `buildBaselineRSFmap` event -- that's the
> "setup section" of my linear script (the part *before* the
> `for (thisYear in years)` loop), translated to use `sim$X` instead
> of bare variables and `P(sim)$X` instead of `params$X`.
>
> Don't write the `simLayers` / `simRSFmap` events yet.

What you should check in the AI's response:

-   Every `sim$X` that's *read* corresponds to a declared
    `expectsInput`, and every `sim$X` that's *written* corresponds to
    a declared `createsOutput`. Mismatches here are the most common
    source of "object not found" errors at run time.
-   Every parameter accessor is `P(sim)$X` (or `Par$X` in newer
    SpaDES.core) -- not `params$X`, not a hard-coded literal. If the
    AI inlined `10` for `numBins`, push back: "use `P(sim)$numBins`
    instead".

### Prompt 3 -- add the recurring events

> Now add two new branches to the `switch()` in
> `doEvent.RSFpredict()`: `simLayers` and `simRSFmap`. Each should be
> the body of one iteration of the original `for (thisYear in years)`
> loop
> -- `simLayers` is the part that builds the year's covariate stack,
> `simRSFmap` is the part that predicts + classifies + writes
> GeoTIFFs. Replace `thisYear` with `as.integer(time(sim))` and
> replace the bare lists `simLand` / `simPred` / `simBinMap` with the
> simList versions `sim$simLand[[key]]`, etc.
>
> At the end of each event, reschedule the same event at
> `time(sim) + P(sim)$predictionInterval` so it recurs.
>
> Also: write the GeoTIFFs under
> `file.path(outputPath(sim), paste0(P(sim)$.studyAreaName, "_sims"))`,
> creating that directory if needed.

The check after this prompt is whether the **self-rescheduling**
trailing line is present in both events. Without it the events fire
once and stop; the dynamic forecast becomes a static one and the bug
is silent -- you get *some* years of output, not a wrong-shaped error.

### What you end up with

The three prompts together produce a 200-300 line `RSFpredict.R`
roughly identical to the [JWTurn/RSFpredict](https://github.com/JWTurn/RSFpredict)
module we used in [@sec-rsfForecast]. Drop it under
`<projectPath>/modules/RSFpredict/RSFpredict.R` and put any helper
functions (like `reclassifyCohortData()`) under
`<projectPath>/modules/RSFpredict/R/`.

## Testing the module

The fastest sanity check is a *minimal* `setupProject()` that runs
just enough of the upstream chain to populate the inputs `RSFpredict`
needs (`cohortData`, `pixelGroupMap`, `timeSinceFire`). Two years are
enough to know whether `simLayers` + `simRSFmap` fire correctly:

```{r test}
repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
options(repos = repos)
if (!require("pak")) install.packages("pak")
pak::pak(c("Require", "SpaDES.project"), ask = FALSE)

library(SpaDES.project)

out <- setupProject(
  paths   = list(projectPath = "~/SpaDES_book/RSFpredict_test"),
  modules = c(
    "PredictiveEcology/Biomass_borealDataPrep@main",
    "PredictiveEcology/Biomass_core@main",
    file.path("PredictiveEcology/scfm@development/modules",
              c("scfmDataPrep", "scfmIgnition", "scfmEscape",
                "scfmSpread", "scfmDiagnostics")),
    "JWTurn/RSFpredict@main"
  ),
  params  = list(
    .globals   = list(.studyAreaName = "dehchoN", .useCache = c(".inputObjects", "init")),
    RSFpredict = list(simulationProcess = "dynamic", predictionInterval = 1)
  ),
  times   = list(start = 2020, end = 2022),
  packages = c("LandR", "climateData"),
  studyArea = reproducible::prepInputs(
    url = "https://drive.google.com/file/d/1ma5qRk2NNidLhrQoiLIzd7W5ogeTaH5-/view",
    fun = "terra::vect"),
  model = reproducible::prepInputs(
    url = "https://drive.google.com/file/d/1ILE0WmePubjHiwynSjV_EyetnUhuQIJU/view",
    fun = "readRDS")
)

results <- SpaDES.core::simInitAndSpades2(out)
SpaDES.core::completed(results)
```

`completed(results)` should show every scheduled event having fired:
`buildBaselineRSFmap` once at year 2020, then `simLayers` and
`simRSFmap` at every year in `2020 ... 2022` because we set
`predictionInterval = 1`. If you only see `init` and
`buildBaselineRSFmap`, the recurring events aren't rescheduling
themselves -- which is exactly the regression the AI's self-scheduling
line was supposed to prevent.

## When things break: iterating with AI

Workflow-level errors fall into a few recognisable shapes. Each one
maps to a short follow-up prompt.

### "Object X not supplied"

> When I run `simInitAndSpades2(out)`, it errors with `object
> 'pixelGroupMap' not supplied`. `RSFpredict` declares it as an
> `expectsInput`, and `Biomass_core` declares it as a
> `createsOutput`. What's wrong?

Three usual answers from the AI: (a) `Biomass_core` isn't actually in
the `modules` list (typo or missing); (b) it *is* in the list, but
listed *after* `RSFpredict`, so `init` of `RSFpredict` runs before
`Biomass_core` creates the object -- fix by reordering; (c) the class
declared in `expectsInput` doesn't match what the upstream module
declares in `createsOutput`, so the wiring is rejected. The AI will
usually tell you which.

### "CRS mismatch" / "extents don't match"

> The `simLayers` event errors in `terra::resample` with "CRS does
> not match". Both rasters say EPSG:3978 when I print them. What
> else could it be?

This is almost always slight numerical drift in the proj4 string
between two layers built by different modules. The fix is one line:
`terra::crs(b) <- terra::crs(a)` before the resample. The AI will
suggest this, or point at `reproducible::postProcess(to = a)` as the
SpaDES-idiomatic version.

### "Predictions only run for some years"

> `completed(results)` shows `simLayers` fired at year 2020 but never
> after. Loop should run every 5 years until 2075. What's wrong?

This is the missing self-reschedule line. Specifically: at the end of
the `simLayers` event there must be a
`sim <- scheduleEvent(sim, time(sim) + P(sim)$predictionInterval,
"RSFpredict", "simLayers")`. The AI will spot this immediately if you
paste the event body.

### "I want predictions every 10 years instead of 5"

You don't change the module -- you change a parameter at the
`setupProject()` call site:

```{r param-override}
params = list(RSFpredict = list(predictionInterval = 10))
```

Which is exactly the point of putting the value in `defineParameter()`
in Prompt 1. The AI will remind you of this if you ask "how do I
change the prediction interval"; if you find yourself editing the
module to change the value, that's a hint the parameter wasn't
exposed.

## See also

[Converting habitat model prediction code into a SpaDES module @sec-rsfPredictModule]

[Forecasting wildlife habitat @sec-rsfForecast]

[Module Files and Metadata @sec-moduleFilesAndMetadata]

[Modules and Events @sec-modulesAndEvents]
