---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: true
  message: true
  error: true
  cache: true
---

# Forecasting wildlife habitat (an RSF) with `setupProject` {#sec-rsfForecast}

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

::: {.callout-note appearance="simple"}
*Several minutes (<10 minutes) -- estimated time for 1st time executing*

*16 GB RAM (or more) -- estimated peak RAM for executing*
:::

```{r setup}
#| include: false
#| eval: true
#| cache: false
#| purl: false

## truncate text output
hook_output <- knitr::knit_hooks$get("output")

knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$outlines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})

qmdPath <- normalizePath(getwd(), winslash = "/")
```

So far we have combined forest succession (`Biomass_core`,
`Biomass_regeneration`) with wildfire (`scfm`) in [Forest Succession and
Wildfire @sec-scfmEg]. A common next question is *ecological*: as the
forest grows and burns over the coming decades, **how does wildlife
habitat change?**

In this chapter we answer that for a caribou Resource Selection Function
(RSF). An RSF is a fitted statistical model that relates the relative
probability of an animal *using* a location to the covariates at that
location (land cover, forest age/biomass, time since fire, etc.). The
`RSFpredict` module takes a *previously fitted* RSF `model` and, at each
time step, re-evaluates it against the *simulated* landscape produced by
the forest and fire modules. The result is a forecast of habitat
selection that responds dynamically to succession and disturbance.

This is a real, three-family project:

-   **forest** -- `Biomass_borealDataPrep`, `Biomass_core`,
    `Biomass_regeneration`
-   **fire** -- the `scfm` family (`scfmDataPrep`, `scfmIgnition`,
    `scfmEscape`, `scfmSpread`, `scfmDiagnostics`)
-   **wildlife** -- `RSFpredict`

Notice that these modules live in *different* GitHub accounts. Because
`SpaDES` modules declare their inputs and outputs in metadata (see
[Module Files and Metadata @sec-moduleFilesAndMetadata]), `setupProject`
can weave them into a single coherent workflow even though they were
written by different people, at different times, for different purposes.

## The project

The whole project is described by a single `setupProject()` call. We
will show it in full first, then walk through the pieces that are new
relative to the forest-and-fire example.

::: {.callout-note appearance="simple"}
This example pulls its fitted RSF `model`, study area, and several input
layers from Google Drive or other sources.
:::

```{r rsf}
# get the latest versions of the predictive ecology R packages
repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
options(repos = repos)
if (!require("pak")) install.packages("pak")
pak::pak(c("Require", "SpaDES.project"), ask = FALSE)

library(SpaDES.project)

projPath <- "~/SpaDES_book/RSF_forecast"

out <- setupProject(
  # Restart = TRUE, # use this to restart your Rstudio session in its own project
  # useGit = "yourGitHubName",   # optional: clone modules as git repos under your account
  paths = list(projectPath = projPath),
  options = list(
    # We host several files on both GoogleDrive and Compute Canada. This redirects the GoogleDrive
    #   file requests to Compute Canada so we don't need to log in to GoogleDrive
    reproducible.urlRemap = 
      "https://object-arbutus.cloud.computecanada.ca/predictiveecology/SCANFI_v2/SCANFI_v2_files_on_arbutus.csv",
    # a shared cloud cache folder, so a whole team can re-use cached results
    reproducible.cloudFolderID =
      "https://drive.google.com/drive/folders/199oEp-TVaCyacwqS4PPf3XWMbhPe4YBN?usp=share_link"
  ),
  modules = c(
    # forest simulation
    "PredictiveEcology/Biomass_borealDataPrep@development",
    "PredictiveEcology/Biomass_core@development",
    "PredictiveEcology/Biomass_regeneration@master",
    # fire simulation -- several modules in one git repository
    file.path("PredictiveEcology/scfm@development/modules",
              c("scfmDataPrep", "scfmIgnition", "scfmEscape",
                "scfmSpread", "scfmDiagnostics")),
    # RSF forecasting
    "JWTurn/RSFpredict@main"
  ),
  params = list(
    .globals = list(
      .plots = "png",
      .studyAreaName = "dehchoN",
      .useCache = c(".inputObjects", "init"),
      dataYear = 2020,        # start year for predicted forest growth
      sppEquivCol = "LandR"
    ),
    scfmDataPrep = list(
      targetN = 2000,         # keep low for a demo; 4000 is better for real estimates
      .useParallelFireRegimePolys = FALSE,
      .useCloud = TRUE
    ),
    Biomass_borealDataPrep = list(
      .useCloud = TRUE
    ),
    RSFpredict = list(
      simulationProcess = "dynamic"
    )
  ),
  packages = c(#"RCurl", "XML", 
               # "snow", 
               # "googledrive", 
               "httr2", 
               "terra",
               "gert", "remotes" # "glmmTMB",
               # "PredictiveEcology/reproducible@development",
               # "PredictiveEcology/LandR@development (>=1.2.0.9002)",
               # "PredictiveEcology/SpaDES.core@development"
               ),
  times = list(start = 2020, end = 2075),

  # ----- Spatial Templates for our Study -----

  studyArea = reproducible::prepInputs(
    url = "https://drive.google.com/file/d/1ma5qRk2NNidLhrQoiLIzd7W5ogeTaH5-/view?usp=share_link",
    fun = "terra::vect",
    destinationPath = "inputs"),

  # buffer the study area to avoid edge effects in the simulation
  studyArea_biomassParam = terra::buffer(studyArea, 10000),  # -- used in Biomass_borealDataPrep
  studyAreaCalibration = studyArea_biomassParam,             # -- used in scfmDataPrep

  # ----- raster templates derived from the inputs -----

  # finer than the final RSF resolution, so we can later compute proportions
  rasterToMatch_biomassParam = {
    rtml <- terra::disagg(modelLand[[1]], fact = 2)
    rtml[] <- 1
    terra::mask(rtml, studyArea_biomassParam)
  },
  rasterToMatchCalibration = rasterToMatch_biomassParam,
  rasterToMatch = {
    reproducible::postProcess(rasterToMatch_biomassParam, cropTo = studyArea, maskTo = studyArea)
  },

  ### Other objects
  # the previously-fitted RSF model object -- used in RSFpredict.R
  model = reproducible::prepInputs(
    url = "https://drive.google.com/file/d/1ILE0WmePubjHiwynSjV_EyetnUhuQIJU/view?usp=share_link",
    fun = "readRDS",
    destinationPath = "inputs"),

  # the land cover the RSF covariates were originally extracted from -- used in RSFpredict.R
  modelLand = reproducible::prepInputs(
    url = "https://drive.google.com/file/d/1LeYrZBKrEPq6jSIP1SWM-grfINTeQjvn/view?usp=share_link",
    fun = "terra::rast",
    destinationPath = "inputs"),

  # start with the historic fire history so scfm has a realistic time-since-fire
  timeSinceFire = {
    rsts <- reproducible::prepInputs(
      url = "https://drive.google.com/file/d/1D6_Zc-xOttWzTOqasXh6KpGTaXYZ5OMA/view?usp=share_link",
      fun = "terra::rast",
      destinationPath = "inputs")
    tsf <- reproducible::postProcess(rsts$timeSinceFire, to = rasterToMatch)
    tsf <- terra::subst(tsf, 122, NA)   # scfm needs non-forest as NA
    tsf
  }
)

results <- SpaDES.core::simInitAndSpades2(out)
```

## Walking through the new pieces

The structure is the same `setupProject()` we have used throughout
([SpaDES workflows and projects](setupProject.qmd)): `modules`,
`params`, `packages`, `times`, and a set of named objects. A few aspects
are worth highlighting because they are typical of a *real* project.

### Composing three model families

The `modules` argument lists modules from two different GitHub accounts
(`PredictiveEcology`, and `JWTurn` for `RSFpredict`). The `scfm` fire
model is a *family of modules within a single repository*, so we build
those paths with `file.path()`. We do not have to tell `SpaDES` how
these fit together -- the metadata does that. The forest modules
*create* the initial biomass and age layers from data (currently SCANFI V2.0 for all
age, species, biomass layers), `scfm` *creates* the initial disturbance layers from
SCANFI, and `RSFpredict` *consumes* both as its initial inputs. These 3 
module families than alternate back and forth through time, feeding each other
derived values of these same layers. They are thus, *dynamically* interacting 
through forecasted conditions.

### `model`: a fitted statistical object as an input

The `model` object is a previously-fitted RSF (here loaded from an
`.rds` with `prepInputs(fun = "readRDS")`). This is a good reminder that
a SpaDES **input** can be *any* R object -- not just a map -- including a
fitted model that a module will apply at every time step. The
`RSFpredict` parameter `simulationProcess = "dynamic"` tells the module
to re-predict against the changing landscape rather than once.

#### Why not fit this RSF here?

To fit an RSF, we need the raw data. For this chapter, we do not have permission
to share these data. However, we do have permission to share any RSF models
that came from these data (as long as the model doesn't carry those raw data!). 
With SpaDES, we can use this approach for now, then replace the `model` object
with a SpaDES module that does the fitting. In fact this is a common approach
that we take so that we can refit the statistical model with new data.

### Study areas and raster templates

Real projects almost always carry *several* spatial templates. Here, these are:

-   `studyArea` -- the area we care about. Here it is a
    *specific, fixed* polygon, loaded from a known file and labelled
    `.studyAreaName = "dehchoN"`. That specificity is what makes cloud
    caching pay off (see [Cloud caching for a team] below);
-   `studyArea_biomassParam` -- a buffered version (here 10 km) so that
    neighbourhood processes (like fire spread) are not distorted at the
    edges of the `studyArea`;
-   `studyAreaCalibration` -- the area used to calibrate the fire
    regime (here the same as `studyArea_biomassParam`).

Each comes with a matching raster template (`rasterToMatch`,
`rasterToMatch_biomassParam`, `rasterToMatchCalibration`). Note how these are
*derived* from one another inside the `setupProject()` call: `studyArea`
is fetched, `studyArea_biomassParam` is `terra::buffer()`ed from it, and
`rasterToMatch` is `postProcess()`ed down from `rasterToMatch_biomassParam`.
Objects defined earlier in the call are available to later ones, so the
project description stays in one place.

::: {.callout-tip appearance="simple"}
Building `rasterToMatch_biomassParam` at a *finer* resolution
(`terra::disagg(..., fact = 2)`) than the final RSF grid lets the
workflow aggregate fine-grained results into proportions later -- a
common pattern when the response variable is a proportion of a coarser
cell.
:::

### Cloud caching for a team

Two things here make the cache *shareable*:
`options(reproducible.cloudFolderID = ...)` points at a Google Drive
folder, and the per-module `.useCloud = TRUE` parameters
(`scfmDataPrep`, `Biomass_borealDataPrep`) opt those expensive
data-preparation and fitting steps into it. We don't *need* this: the important
part of a re-runnable, reproducible, re-usable workflow is that the workflow
changes very little or not at all to get these nice behaviours (refit with new
data AND use somebody else's fit if it is already done).

The reason this works is the **specific `studyArea`**. `Cache` decides
whether a result can be reused by digesting a function's inputs (see
[Introduction to Cache @sec-caching]); because the study area here is a
*fixed* polygon rather than something randomly generated, every user who
runs this project feeds those slow modules the *identical* inputs, and
therefore computes the *identical* cache key. So once **one** team member
has run the slow `scfmDataPrep` and `Biomass_borealDataPrep` fitting and
written the result to the shared cloud folder, everyone else *downloads*
that cached result instead of re-running it.

Contrast this with the forest-and-fire example, which builds its study
area with `LandR::randomStudyArea()`: a randomly-drawn area would change
the inputs (and the cache key) from user to user, so the cloud cache
could not be re-used. Pinning a specific study area is what turns the
cloud cache into a genuine team -- or workshop or beyond -- resource.

### `simInitAndSpades2()`

We launch the run with `SpaDES.core::simInitAndSpades2(out)`. The `2`
variants of `simInit2()` / `simInitAndSpades2()` simply take the *list*
returned by `setupProject()` directly, instead of having to
`do.call()` it. If the run is interrupted, `SpaDES.core::restartSpades()`
can resume it.

## Examining the forecast

Once the run finishes, the same accessors used elsewhere apply. Because
we set `.plots = "png"`, every module that plots with the `Plots`
function will have written figures into the `outputs/figures` folder --
including the forecasted RSF surface through time.

```{r examine}
completed(results)                       # the events that ran
elapsedTime(results, units = "mins")     # time per event

# the forecasted RSF prediction(s) created by RSFpredict
# (object name depends on the module's metadata)
moduleMetadata(results)$RSFpredict$outputObjects
```

## See also

[Forest Succession and Wildfire @sec-scfmEg]

[SpaDES workflows and projects](setupProject.qmd)

[`reproducible::prepInputs` for Data @sec-prepInputs]

[Introduction to Cache @sec-caching]

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path(qmdPath, "R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
