---
author: "Eliot McIntire and Julie Turner"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# Converting habitat model prediction code into a SpaDES module {#sec-rsfPredictModule}

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

::: {.callout-tip appearance="simple"}
See the [Glossary](glossary.qmd) for definitions of recurring terms
(e.g. *SpaDES module*, *SpaDES event*, *continuous workflow*,
*interoperable model*).
:::

In the previous chapter, we built up a `setupProject()` call from a
hand-rolled script. That moved the **project-level bookkeeping**
(packages, paths, modules, inputs) out of the script and into one
declarative call. This chapter takes the next step: turning the
**ecology** -- the actual prediction code -- into a `SpaDES` module so
it can be composed with the forest (`LandR`) and fire (`scfm`) families
into the dynamic forecast we ran in [Forecasting wildlife habitat
@sec-rsfForecast].

The starting point is a linear R script that runs an RSF forecast on
its own: one setup section (load inputs, build the baseline map) and
one annual loop (re-evaluate the model against simulated landscapes,
write GeoTIFFs). The end state is `RSFpredict.R` -- the same logic,
factored into a `SpaDES` module's three required pieces:

1.  **Metadata** -- a `defineModule()` block that declares what the
    module needs, what it produces, and what its parameters are.
2.  **Init event** -- runs once at the start of the simulation; it
    builds the baseline map and **schedules** the recurring annual
    events.
3.  **Recurring events** -- the body of the original annual loop,
    split into `simLayers` (build the year's land covariates) and
    `simRSFmap` (predict + classify + write).

We will build the module up in that order.

## Bootstrap

```{r bootstrap}
#| eval: true
repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
options(repos = repos)
if (!require("pak")) install.packages("pak")
pak::pak(c("Require", "SpaDES.project"), ask = FALSE)

library(SpaDES.project)
```

## The starting point: a plain R script

Assume the inputs (`model`, `studyArea`, `modelLand`, `cohortData`,
`pixelGroupMap`, `timeSinceFire`, `rasterToMatch`) are already in the
workspace and a helper `reclassifyCohortData()` is sourced. The script
runs one setup section then a year-by-year loop:

```{r linearScript}
#| code-fold: true
#| code-summary: "Show full linear script"

## ---- parameters ------------------------------------------------------------
params <- list(
  numBins            = 10L,
  simulationProcess  = "dynamic",
  predictionInterval = 5,
  predictStartYear   = 2025,
  predictLastYear    = TRUE,
  ts_else            = 100L,
  studyAreaName      = "dehchoN"
)

outDir <- reproducible::checkPath(
  file.path("outputs", paste0(params$studyAreaName, "_sims")), create = TRUE)

startYear <- 2020
endYear   <- 2075

## ---- SETUP: build the baseline RSF map -------------------------------------
modLand <- modelLand
modLand$timeSinceFire[is.na(modLand$timeSinceFire)] <- params$ts_else
if (!is.null(weights(model))) {
  w <- modelLand[[1]]; w[] <- 1; names(w) <- "w"
  modLand <- c(modLand, w)
}
pred      <- terra::predict(modLand, model, type = "response", re.form = NA)
pred.mask <- terra::mask(pred, studyArea)
breaks    <- terra::global(pred.mask, quantile, na.rm = TRUE,
                           probs = seq(0, 1, 1 / params$numBins))
binMap    <- terra::classify(pred.mask, t(breaks),
                             include.lowest = TRUE, brackets = TRUE)

## ---- ANNUAL LOOP -----------------------------------------------------------
simLand   <- list(); simPred <- list(); simBinMap <- list()
years <- seq(params$predictStartYear, endYear, by = params$predictionInterval)
if (params$predictLastYear && tail(years, 1L) != endYear) years <- c(years, endYear)

for (thisYear in years) {
  key <- paste0("year", thisYear)
  template       <- rasterToMatch
  templateCoarse <- terra::aggregate(template, fact = 2)

  reclassForest <- reclassifyCohortData(
    cohortData = cohortData, sppEquivCol = "LandR",
    pixelGroupMap = pixelGroupMap, mixedForestCutoffs = c(0.33, 0.66))
  ft <- reclassForest$`forest type`

  needle_mask <- terra::classify(ft,
    rcl = matrix(c(210, 1), ncol = 2, byrow = TRUE), others = 0)
  prop_needleleaf <- terra::resample(needle_mask, template, method = "average")
  names(prop_needleleaf) <- "prop_needleleaf"

  mixed_mask <- terra::classify(ft,
    rcl = matrix(c(220, 1, 230, 1), ncol = 2, byrow = TRUE), others = 0)
  prop_mixedforest <- terra::resample(mixed_mask, template, method = "average")
  names(prop_mixedforest) <- "prop_mixedforest"

  forests <- reproducible::postProcess(
    c(prop_needleleaf, prop_mixedforest), templateCoarse, method = "average")

  tsf <- reproducible::postProcess(timeSinceFire, to = templateCoarse)
  tsf[is.na(tsf)] <- params$ts_else + (thisYear - 2020)
  names(tsf) <- "timeSinceFire"

  simLand[[key]] <- c(forests, tsf)
  terra::writeRaster(simLand[[key]],
    file.path(outDir, paste0("mapLayers_", params$studyAreaName, "_", key, ".tif")),
    overwrite = TRUE)

  modLand <- simLand[[key]]
  if (!is.null(weights(model))) {
    w <- modLand[[1]]; w[] <- 1; names(w) <- "w"
    modLand <- c(modLand, w)
  }
  simPred[[key]]   <- terra::predict(modLand, model, type = "response", re.form = NA)
  pred.mask        <- terra::mask(simPred[[key]], studyArea)
  breaks           <- terra::global(pred.mask, quantile, na.rm = TRUE,
                                    probs = seq(0, 1, 1 / params$numBins))
  simBinMap[[key]] <- terra::classify(pred.mask, t(breaks),
                                      include.lowest = TRUE, brackets = TRUE)

  terra::writeRaster(pred.mask,
    file.path(outDir, paste0("rsfMap",    params$studyAreaName, "_", key, ".tif")),
    overwrite = TRUE)
  terra::writeRaster(simBinMap[[key]],
    file.path(outDir, paste0("rsfBinMap_", params$studyAreaName, "_", key, ".tif")),
    overwrite = TRUE)
}
```

This script works -- on the machine where it was written, and only as
long as someone else loads exactly the right inputs into exactly the
right variable names beforehand. It is also impossible to compose with
another module's outputs in a principled way: `cohortData`,
`pixelGroupMap`, `timeSinceFire` have to *arrive* somehow. As a module,
the same code declares those needs and gets them filled in by whoever
runs the project.

## Step 0: scaffold the module with `newModule()`

Before writing anything, generate the module skeleton with
`SpaDES.core::newModule()`. It creates the directory layout, the main
`.R` file with stubbed `defineModule()` / `doEvent.RSFpredict()` /
`.inputObjects()`, an empty `R/` folder for helper functions, plus
`README.md`, `NEWS.md`, `RSFpredict.Rmd`, and `citation.bib`. We then
*edit* the stub rather than writing the whole file from scratch.

```{r newModule}
SpaDES.core::newModule(
  name       = "RSFpredict",
  path       = "~/SpaDES_book/RSF_forecast/modules",   # parent dir
  open       = FALSE                                    # don't auto-open in RStudio
)
```

This drops the following files into
`~/SpaDES_book/RSF_forecast/modules/RSFpredict/`:

```
RSFpredict/
|-- RSFpredict.R       <- defineModule + doEvent.RSFpredict + .inputObjects (stubs)
|-- RSFpredict.Rmd     <- documentation template
|-- README.md
|-- NEWS.md
|-- citation.bib
|-- LICENSE.md
|-- data/
|-- tests/
\-- R/                 <- put helper functions here (e.g. reclassifyCohortData.R)
```

Open `RSFpredict.R` -- it already contains the *shape* we are aiming
for, with `# ! ----- EDIT BELOW ----- !` markers showing where to put
each section. The next three steps replace the stub contents with the
ecology from our linear script.

## Step 1: fill in the metadata

The scaffold's `defineModule()` block already has the right shape --
it's just stuffed with placeholders (empty `reqdPkgs`, a single example
`defineParameter`, empty `inputObjects`/`outputObjects`). Replace those
with the real ones. Three pieces matter:

-   **`reqdPkgs`** -- what the module needs to be installed
    (`setupProject()` reads this and reconciles it with what the rest
    of the project needs);
-   **`parameters`** -- every knob the script exposed (`numBins`,
    `simulationProcess`, etc.), with type + default;
-   **`inputObjects` / `outputObjects`** -- the names + classes the
    module *expects* to find in the simList, and the names + classes it
    *produces*. This is what lets `setupProject()` and the other
    modules wire themselves together.

```{r meta}
defineModule(sim, list(
  name        = "RSFpredict",
  description = "Forecast a fitted RSF against simulated landscapes",
  keywords    = c("RSF", "habitat", "caribou"),
  authors     = structure(list(list(given = c("Julie", "W"),
                                    family = "Turner",
                                    role = c("aut", "cre"))),
                          class = "person"),
  version     = list(RSFpredict = "0.0.0.9000"),
  timeunit    = "year",
  documentation = list("README.md"),

  ## packages the module needs -- merged with whoever else's at setupProject()
  reqdPkgs = list(
    "reproducible", "terra", "ggplot2", "glmmTMB",
    "tidyterra", "viridis", "ggthemes",
    "SpaDES.core (>= 3.0.3.9003)"
  ),

  ## every `params$X` in the linear script becomes a defineParameter()
  parameters = bindrows(
    defineParameter("numBins",            "integer",   10L,        NA, NA,
                    "Number of bins for the classified RSF map"),
    defineParameter("simulationProcess",  "character", "static",   NA, NA,
                    "Should the simulation use LandR ('dynamic') or fixed cover ('static')?"),
    defineParameter("predictionInterval", "numeric",   5,          NA, NA,
                    "Years between predictions"),
    defineParameter("predictStartYear",   "numeric",   2025,       NA, NA,
                    "First year to forecast"),
    defineParameter("predictLastYear",    "logical",   TRUE,       NA, NA,
                    "Force a prediction at the final simulation year"),
    defineParameter("ts_else",            "integer",   100L,       NA, NA,
                    "Fill value for NA cells in time-since-fire"),
    defineParameter(".studyAreaName",     "character", NA_character_, NA, NA,
                    "Human-readable name for the study area")
  ),

  ## inputs the linear script assumed were already in scope
  inputObjects = bindrows(
    expectsInput("model",         "glmmTMB",    "Fitted RSF model"),
    expectsInput("studyArea",     "SpatVector", "Unbuffered study area"),
    expectsInput("modelLand",     "SpatRaster", "Cover layers the model was fit on"),
    expectsInput("cohortData",    "data.table", "LandR cohort data"),
    expectsInput("pixelGroupMap", "SpatRaster", "LandR pixel-group raster"),
    expectsInput("timeSinceFire", "SpatRaster", "Time-since-fire raster (from scfm)"),
    expectsInput("rasterToMatch", "SpatRaster", "Spatial template")
  ),

  ## things the linear script wrote to `pred`, `binMap`, `simLand`, ...
  outputObjects = bindrows(
    createsOutput("pred",      "SpatRaster", "Baseline RSF prediction"),
    createsOutput("binMap",    "SpatRaster", "Baseline RSF prediction binned"),
    createsOutput("simLand",   "list",       "Per-year stack of land covariates"),
    createsOutput("simPred",   "list",       "Per-year RSF predictions"),
    createsOutput("simBinMap", "list",       "Per-year binned RSF predictions")
  )
))
```

Two practical effects of writing the metadata first:

-   `setupProject()` now knows it has to install `glmmTMB`, `tidyterra`,
    and the rest before this module is asked to run. The user doesn't
    list those in their `packages` argument -- the module brought them.
-   The forest and fire modules can be plugged in *because the names
    match*: `LandR::Biomass_core` produces `cohortData` and
    `pixelGroupMap`; `scfm` produces `timeSinceFire`. Their
    `createsOutput()` declarations meet our `expectsInput()`
    declarations and `SpaDES` connects them.

## Step 2: fill in the `init` event

`newModule()` left you a stub `doEvent.RSFpredict()` -- the `switch()`
shell with `init`, `plot`, and `save` placeholder branches. We keep
the `switch()` and replace what's inside the `init` branch. The
**`init` event** runs exactly once, at the start of the simulation,
and is where the linear script's *setup section* lives -- but with
one extra job: telling `SpaDES` what to do next.

We rename the setup section's "build the baseline" block to
`buildBaselineRSFmap` (an event), and `init` becomes a tiny scheduler:

```{r init}
doEvent.RSFpredict <- function(sim, eventTime, eventType) {
  switch(
    eventType,

    init = {
      ## run once: build the baseline RSF map immediately
      sim <- scheduleEvent(sim, time(sim),
                           "RSFpredict", "buildBaselineRSFmap")

      ## if dynamic, also start the recurring loop at predictStartYear
      if (P(sim)$simulationProcess == "dynamic") {
        sim <- scheduleEvent(sim, P(sim)$predictStartYear,
                             "RSFpredict", "simLayers")
        sim <- scheduleEvent(sim, P(sim)$predictStartYear,
                             "RSFpredict", "simRSFmap")
      }

      ## optionally pin a final-year prediction even if it doesn't fall
      ## on a predictionInterval step
      if (P(sim)$predictLastYear) {
        sim <- scheduleEvent(sim, end(sim), "RSFpredict", "simLayers")
      }
    },

    buildBaselineRSFmap = {
      ## the *content* is the linear setup section, with sim$ in place
      ## of bare variables
      modLand <- sim$modelLand
      modLand$timeSinceFire[is.na(modLand$timeSinceFire)] <- P(sim)$ts_else
      if (!is.null(weights(sim$model))) {
        w <- sim$modelLand[[1]]; w[] <- 1; names(w) <- "w"
        modLand <- c(modLand, w)
      }
      sim$pred  <- terra::predict(modLand, sim$model,
                                  type = "response", re.form = NA)
      pred.mask <- terra::mask(sim$pred, sim$studyArea)
      breaks    <- terra::global(pred.mask, quantile, na.rm = TRUE,
                                 probs = seq(0, 1, 1 / P(sim)$numBins))
      sim$binMap <- terra::classify(pred.mask, t(breaks),
                                    include.lowest = TRUE, brackets = TRUE)
    }

    ## (the recurring events go here -- next section)
  )
  return(invisible(sim))
}
```

Three things changed translating the setup section into the module:

-   **Plain variables become `sim$X`.** The simList is the module's
    scope; instead of `model`, `studyArea`, etc. coming from the caller's
    workspace, they live on the simList and the module reads/writes
    them as `sim$model`, `sim$pred`, ....
-   **Parameters change accessor.** `params$numBins` becomes
    `P(sim)$numBins`. Same value, but the source is now the simList,
    which means a user can override defaults via
    `setupProject(params = list(RSFpredict = list(numBins = 20)))`
    without editing the module.
-   **The loop is gone (for now).** Instead of iterating in R,
    `init` *schedules* `simLayers` and `simRSFmap` at the right times
    and `SpaDES` will fire them. This is what makes the module
    interleave correctly with other modules: at each timestep, the
    scheduler can call `Biomass_core` to advance the forest, then
    `scfm` to advance fire, then our `simLayers` to assemble those
    outputs into RSF covariates, then our `simRSFmap` to predict.

## Step 3: add the recurring events

`simLayers` and `simRSFmap` aren't in the scaffold -- they're the
event names you invented for this module. Add them as new `switch()`
branches alongside `init`. They are the body of the original annual
loop, split at the natural seam: build inputs, then predict from them.
Each ends by scheduling **itself** at `time(sim) + predictionInterval`,
so they recur:

```{r recurring}
    simLayers = {
      thisYear <- as.integer(time(sim))
      key      <- paste0("year", thisYear)

      template       <- sim$rasterToMatch
      templateCoarse <- terra::aggregate(template, fact = 2)

      reclassForest <- reclassifyCohortData(
        cohortData         = sim$cohortData,
        sppEquivCol        = "LandR",
        pixelGroupMap      = sim$pixelGroupMap,
        mixedForestCutoffs = c(0.33, 0.66)
      )
      ft <- reclassForest$`forest type`

      needle_mask <- terra::classify(ft,
        rcl = matrix(c(210, 1), ncol = 2, byrow = TRUE), others = 0)
      prop_needleleaf <- terra::resample(needle_mask, template,
                                         method = "average")
      names(prop_needleleaf) <- "prop_needleleaf"

      mixed_mask <- terra::classify(ft,
        rcl = matrix(c(220, 1, 230, 1), ncol = 2, byrow = TRUE), others = 0)
      prop_mixedforest <- terra::resample(mixed_mask, template,
                                          method = "average")
      names(prop_mixedforest) <- "prop_mixedforest"

      forests <- reproducible::postProcess(
        c(prop_needleleaf, prop_mixedforest),
        templateCoarse, method = "average")

      tsf <- reproducible::postProcess(sim$timeSinceFire, to = templateCoarse)
      tsf[is.na(tsf)] <- P(sim)$ts_else + (thisYear - 2020)
      names(tsf) <- "timeSinceFire"

      sim$simLand[[key]] <- c(forests, tsf)

      outDir <- reproducible::checkPath(
        file.path(outputPath(sim),
                  paste0(P(sim)$.studyAreaName, "_sims")), create = TRUE)
      terra::writeRaster(
        sim$simLand[[key]],
        file.path(outDir, paste0("mapLayers_",
                                 P(sim)$.studyAreaName, "_", key, ".tif")),
        overwrite = TRUE)

      ## schedule self
      sim <- scheduleEvent(sim, time(sim) + P(sim)$predictionInterval,
                           "RSFpredict", "simLayers")
    },

    simRSFmap = {
      thisYear <- as.integer(time(sim))
      key      <- paste0("year", thisYear)

      modLand <- sim$simLand[[key]]
      if (!is.null(weights(sim$model))) {
        w <- modLand[[1]]; w[] <- 1; names(w) <- "w"
        modLand <- c(modLand, w)
      }
      sim$simPred[[key]] <- terra::predict(modLand, sim$model,
                                           type = "response", re.form = NA)
      pred.mask <- terra::mask(sim$simPred[[key]], sim$studyArea)
      breaks    <- terra::global(pred.mask, quantile, na.rm = TRUE,
                                 probs = seq(0, 1, 1 / P(sim)$numBins))
      sim$simBinMap[[key]] <- terra::classify(pred.mask, t(breaks),
                                              include.lowest = TRUE,
                                              brackets = TRUE)

      outDir <- reproducible::checkPath(
        file.path(outputPath(sim),
                  paste0(P(sim)$.studyAreaName, "_sims")), create = TRUE)
      terra::writeRaster(
        pred.mask,
        file.path(outDir, paste0("rsfMap",
                                 P(sim)$.studyAreaName, "_", key, ".tif")),
        overwrite = TRUE)
      terra::writeRaster(
        sim$simBinMap[[key]],
        file.path(outDir, paste0("rsfBinMap_",
                                 P(sim)$.studyAreaName, "_", key, ".tif")),
        overwrite = TRUE)

      ## schedule self
      sim <- scheduleEvent(sim, time(sim) + P(sim)$predictionInterval,
                           "RSFpredict", "simRSFmap")
    },
```

What's different from the linear loop:

-   **No `for (thisYear in years)`.** The "current year" is just
    `time(sim)`, and the recurrence is the trailing `scheduleEvent()`
    call. `SpaDES` walks the queue and fires our events at the right
    moments -- *interleaved* with whatever other modules' events fall
    between.
-   **`simLayers` and `simRSFmap` are separate events**, even though
    they always run for the same year. That separation is what lets
    *other* modules (the forest and fire ones) update `cohortData` /
    `timeSinceFire` **between** `simLayers` of year 2025 and
    `simLayers` of year 2030.
-   **Outputs accumulate on the simList**, not in standalone
    `list()`s. `sim$simLand[[key]] <- ...` is the module's way of
    saying "here is this year's covariate stack; anyone downstream can
    use it".

## Running it

Once the module exists on disk (with `RSFpredict.R` plus the
`reclassifyCohortData()` helper under `R/`), it slots into the same
`setupProject()` call we used in [Forecasting wildlife habitat
@sec-rsfForecast]:

```{r run}
out <- setupProject(
  paths    = list(projectPath = "~/SpaDES_book/RSF_forecast"),
  modules  = c(
    "PredictiveEcology/Biomass_borealDataPrep@development",
    "PredictiveEcology/Biomass_core@development",
    "PredictiveEcology/Biomass_regeneration@master",
    file.path("PredictiveEcology/scfm@development/modules",
              c("scfmDataPrep", "scfmIgnition", "scfmEscape",
                "scfmSpread", "scfmDiagnostics")),
    "JWTurn/RSFpredict@main"
  ),
  params = list(
    RSFpredict = list(simulationProcess = "dynamic")
  ),
  times = list(start = 2020, end = 2075),
  ## ... and your studyArea / model / modelLand prepInputs as before
)

results <- SpaDES.core::simInitAndSpades2(out)
```

## What the conversion bought us

Same code, three new capabilities:

-   **Composability.** Because the `inputObjects` declarations name
    `cohortData`, `pixelGroupMap`, and `timeSinceFire`, the forest and
    fire modules can supply them. The linear script had no way to ask
    for those.
-   **Scheduling.** The `simulationProcess = "dynamic"` parameter
    *flips a branch* in `init`. The same module supports static and
    dynamic modes without two scripts.
-   **Reproducibility.** `Cache()` calls on the expensive
    `terra::predict` step memoize across runs; metadata + setupProject
    means the same `out` produces the same results on any machine.

## See also

[Forecasting wildlife habitat @sec-rsfForecast]

[Module Files and Metadata @sec-moduleFilesAndMetadata]

[Modules and Events @sec-modulesAndEvents]

[Scheduling @sec-scheduling]

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path("R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
