---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# Saving in SpaDES {#sec-saving}

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

A reproducible workflow eventually needs to put things *on disk*: maps,
tables, fitted models, or the entire state of a simulation so that it
can be revisited, shared, or post-processed. Like inputs (see [Providing
Module Inputs @sec-moduleinputs]), `SpaDES` treats saving as a
first-class, declarative concept: instead of scattering `saveRDS()`
calls through your code, you describe *what* to save and *when*, and
`SpaDES` schedules the saving for you.

There are three levels at which saving can be controlled, from most to
least modular:

| Level | Mechanism | Who usually sets it |
|-------|-----------|---------------------|
| **Model** | the `outputs` argument / `outputs(sim)` | the *user* of a set of modules |
| **Module** | `.saveObjects` parameter + `saveFiles()` | the module *developer* |
| **Custom** | `saveRDS()`, `qs2::qs_save()`, `terra::writeRaster()`, etc. | the module *developer* |

A fourth, distinct mechanism saves the *whole* `simList` (see
[Saving and reloading a whole `simList`]).

::: {.callout-note appearance="simple"}
Most of the machinery below already exists and is documented in
`SpaDES.core`. This chapter is a practical orientation; for the
authoritative reference see the `SpaDES.core` *modules* vignette
([`vignette("ii-modules", package = "SpaDES.core")`](https://rdrr.io/cran/SpaDES.core/f/vignettes/ii-modules.Rmd))
and `?SpaDES.core::saveFiles`.
:::

```{r pkgs}
#| eval: true

repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
options(repos = repos)
if (!require("pak")) install.packages("pak")
pak::pak(c("Require", "SpaDES.project"), ask = FALSE)

library(SpaDES.project)
out <- setupProject(
  packages = c("SpaDES.core", "terra"),
  options = list(repos = repos),
  paths = list(projectPath = "~/SpaDES_book/Saving")
)
```

## Model-level saving: the `outputs` argument

The most common way to save is to hand `simInit()` (or
`simInitAndSpades()`) an `outputs` `data.frame`. It is the saving
counterpart of the `inputs` argument: each row names an object in the
`simList` and, optionally, when and how to save it.

The `data.frame` needs, minimally, a column named `objectName`. The
other columns are filled in with sensible defaults if absent:

| Column | Description |
|--------|-------------|
| **`objectName`** | Name of the object in the `simList` to save. |
| **`file`** | Base name for the saved file. Defaults to `objectName`. |
| **`fun`** | Function used to save. Defaults to `saveRDS`. |
| **`package`** | Package providing `fun`. |
| **`saveTime`** | Simulation time at which to save. Defaults to `end(sim)`. |
| **`arguments`** | A (list of) list(s) of extra arguments passed to `fun`. |

`SpaDES` adds a seventh column, `saved` (logical), which flips to `TRUE`
once the save has happened, so `outputs(sim)` doubles as a record of
what has already been written to disk.

In the simplest case, we just name the object. The file lands in the
simulation's `outputPath`, as an `.rds`, at the end of the run:

```{r outputs-simple}
outputDir <- "~/SpaDES_book/Saving/outputs"

mySim <- simInit(
  times = list(start = 0, end = 5),
  modules = "randomLandscapes",
  params = list(.globals = list(stackName = "landscape")),
  paths = list(modulePath = getSampleModules(tempdir()),
               outputPath = outputDir),
  outputs = data.frame(objectName = "landscape")
)

outputs(mySim)        # one row, saved = NA (not yet saved)

mySim <- spades(mySim)

outputs(mySim)        # saved is now TRUE; note the file path
```

To save the *same* object several times, give it several rows with
different `saveTime`s. To control the format, set `fun`, `package`, and
`arguments`. Here we write a `SpatRaster` as a GeoTIFF at times 3 and 6.
Note that we can set `outputs()` on an existing `simList` with the
replacement function, rather than only at `simInit()`:

```{r outputs-fun}
outputs(mySim) <- data.frame(
  objectName = "landscape",
  fun        = "writeRaster",
  package    = "terra",
  saveTime   = c(3, 6),
  arguments  = I(lapply(c(3, 6), function(x) {
    list(datatype = "FLT4S", filetype = "GTiff", overwrite = TRUE)
  }))
)

mySim <- spades(mySim)
outputs(mySim)
```

::: {.callout-tip appearance="simple"}
Because `outputs` is "model-level", a *user* can decide what to save
without editing any module code. This is the recommended starting point.
:::

## Module-level saving: `.saveObjects` and `saveFiles()`

A module *developer* can build saving into a module so that future users
only need to toggle a few parameters. This relies on a set of `.save*`
parameters and the `saveFiles()` helper. When a module is created with
`newModule()`, a `save` event already exists; it just needs to be
scheduled.

| Parameter | Description |
|-----------|-------------|
| **`.saveInitialTime`** | Time of the first save (between `start` and `end`). `NA` means no saving. |
| **`.saveInterval`** | Interval between repeated saves. |
| **`.saveObjects`** | Character vector of object names to save. |
| **`.savePath`** | Where to save (defaults to the `outputPath`). |

```{r save-objects}
mySim <- simInit(
  times = list(start = 0, end = 5),
  modules = "randomLandscapes",
  params = list(
    .globals = list(stackName = "landscape"),
    randomLandscapes = list(
      .saveInitialTime = 0, .saveInterval = 2,
      .saveObjects = "landscape",
      .savePath = file.path(outputDir, "randomLandscapes")
    )
  ),
  paths = list(modulePath = getSampleModules(tempdir()),
               outputPath = outputDir)
)

mySim <- spades(mySim)
```

Inside a module, the save itself is performed by `saveFiles(sim)`, and a
recurring save is kept going by re-scheduling the event:

```{r save-events}
### WITHIN A MODULE event:
sim <- saveFiles(sim)

# reschedule the next save
nextSave <- time(sim) + P(sim)$.saveInterval
sim <- scheduleEvent(sim, nextSave, currentModule(sim), "save")
```

See [Scheduling Events @sec-schedulingEvents] for more on
`scheduleEvent()`.

## Custom saving inside a module

Sometimes the cleanest thing is to save directly with a normal R
function -- `saveRDS()`, `qs2::qs_save()`, `terra::writeRaster()`,
`data.table::fwrite()`, and so on -- from inside an event. This is the
*least* modular approach (it happens whether the user wants it or not),
but it is sometimes the most practical.

When you save this way, it is good practice to *register* the file with
the `simList` using `registerOutputs()`, so that the file is tracked in
`outputs(sim)` and is picked up by `saveSimList()` (below):

```{r custom-save}
### WITHIN A MODULE event:
fname <- file.path(outputPath(sim), paste0("predictions_", time(sim), ".rds"))
saveRDS(sim$predictions, fname)

# record that we saved it
sim <- registerOutputs(fname, sim)
```

## Saving and reloading a whole `simList`

Saving individual objects is not the same as saving the *simulation*. A
`simList` cannot be reliably saved with `save()` or `saveRDS()` for two
reasons:

1.  modules rely on **active bindings** (these are what make `mod` and
    `P(sim)`/`Par` work); and
2.  file-backed objects such as `terra` `SpatRaster`s live partly on
    disk, not in the R object.

`SpaDES.core` provides `saveSimList()` and `loadSimList()` to handle
both correctly:

```{r saveSimList}
simFile <- "~/SpaDES_book/Saving/mySim.qs2"

saveSimList(mySim, filename = simFile)

# later, possibly on another machine
mySim2 <- loadSimList(simFile)
```

If there are file-backed objects (or you ask to include the `outputs`
and `inputs` files), `saveSimList()` transparently writes an *archive*
(`.tar.gz`, or `.zip` on Windows) so everything travels together:

```{r saveSimList-archive}
saveSimList(mySim, filename = simFile, outputs = TRUE, inputs = TRUE)

# reload an archived simList
mySim2 <- unzipSimList(sub("\\.qs2$", ".tar.gz", simFile))
```

::: {.callout-note appearance="simple"}
`saveSimList()` is the right tool for *checkpointing*, *sharing*, or
*archiving* a run. To merely speed up re-running, prefer caching instead
-- see [Introduction to Cache @sec-caching].
:::

## Where do saved files go?

Unless a `file` or `.savePath` says otherwise, saved objects land in the
simulation's `outputPath`:

```{r outputPath}
outputPath(mySim)
```

The output path is set via the `paths` argument to `setupProject()`,
`simInit()`, or `simInitAndSpades()` (the `outputPath` element), in the
same way the cache path is set (see [Caching @sec-caching]).

## Best practices

1.  Prefer **model-level** `outputs` for things a *user* chooses to
    save; it requires no module edits.
2.  Build **module-level** saving (`.saveObjects`) into modules whose
    users will predictably want certain objects.
3.  When you save with a custom function, **`registerOutputs()`** the
    file so it is tracked.
4.  Use **`saveSimList()`/`loadSimList()`** to checkpoint or share an
    entire run -- not `saveRDS()`.
5.  Save only what you need: large file-backed objects can make archives
    very large.

## See also

[`vignette("ii-modules", package = "SpaDES.core")`](https://rdrr.io/cran/SpaDES.core/f/vignettes/ii-modules.Rmd)

[`?SpaDES.core::saveFiles`](https://rdrr.io/cran/SpaDES.core/man/saveFiles.html)

[`?SpaDES.core::outputs`](https://rdrr.io/cran/SpaDES.core/man/simList-accessors-outputs.html)

[`?SpaDES.core::saveSimList`](https://rdrr.io/cran/SpaDES.core/man/saveSimList.html)

[`?SpaDES.core::loadSimList`](https://rdrr.io/cran/SpaDES.core/man/loadSimList.html)

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path("R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
