---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# `setupProject` for nimble workflows {#sec-setupProjectNimble}

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

Most SpaDES workflows begin life as an ordinary R script: load some
packages, set some paths, download a few things, read a shapefile,
initialize a simulation, run it. That script *works* -- on the machine
where it was written. Moving it to another computer (a colleague's, a
cluster, the version of your own machine six months from now) is where
the trouble starts: a missing package, a hard-coded path that doesn't
exist, a data file that was never in the repository.

This chapter takes one such script and translates it, step by step, into
the equivalent `SpaDES.project::setupProject()` call. The simulation it
runs is identical. What changes is that the *nimble* version is
reproducible and portable by construction: everything it needs is
described declaratively, packages are reconciled in one place, and every
input **starts from a URL, not a local file path**.

## The starting point: a hand-rolled script

Here is a typical "works on my machine" setup for a small `LandR`
Biomass run. It is not wrong -- it is just doing a lot of bookkeeping by
hand, and most of that bookkeeping is exactly what breaks elsewhere.

```{r before}
## 1) Packages -------------------------------------------------------
library(SpaDES.core)
library(SpaDES.project)
library(terra)
library(sf)

## 2) Paths ----------------------------------------------------------
setPaths(
  modulePath  = "~",
  inputPath   = "~",
  outputPath  = "~",
  cachePath   = "~",
  scratchPath = "~"
)

## clean cache to avoid previous mismatches
unlink(getPaths()$cachePath, recursive = TRUE)
dir.create(getPaths()$cachePath, showWarnings = FALSE)

## 3) Download modules ----------------------------------------------
SpaDES.project::getModule(
  modules = c(
    "PredictiveEcology/Biomass_borealDataPrep",
    "PredictiveEcology/Biomass_speciesData",
    "PredictiveEcology/Biomass_core"
  ),
  modulePath = getPaths()$modulePath,
  overwrite  = TRUE
)

## 4) Download study area from Google Drive -------------------------
shapeZipID <- "1gycTXyZzgXIUGM6dyAJbmFxq8TOAm-Ik"
shapeDir   <- file.path(getPaths()$inputPath, "studyArea")
dir.create(shapeDir, recursive = TRUE, showWarnings = FALSE)
shapeZip   <- file.path(shapeDir, "studyArea.zip")

if (!file.exists(shapeZip)) {
  download.file(
    url      = paste0("https://drive.google.com/uc?export=download&id=", shapeZipID),
    destfile = shapeZip,
    mode     = "wb"
  )
}

## unzip ------------------------------------------------------------
if (length(list.files(shapeDir, pattern = "\\.shp$", recursive = TRUE)) == 0) {
  unzip(shapeZip, exdir = shapeDir)
}

## read shapefile ---------------------------------------------------
shpFile <- list.files(shapeDir, pattern = "\\.shp$",
                      recursive = TRUE, full.names = TRUE)
stopifnot(length(shpFile) == 1)

studyArea <- terra::vect(shpFile)
studyArea <- terra::project(studyArea, "EPSG:3978")

## 5) Simulation settings -------------------------------------------
times   <- list(start = 1, end = 1)
modules <- c("Biomass_borealDataPrep", "Biomass_speciesData", "Biomass_core")
objects <- list(studyArea = studyArea, studyAreaLarge = studyArea)

## 6) Initialize and run --------------------------------------------
sim <- simInit(times = times, modules = modules,
               objects = objects, paths = getPaths())
sim <- spades(sim)
```

## The same workflow with `setupProject`

The same run, expressed declaratively:

```{r after}
if (!require("pak")) install.packages("pak")
pak::pak(c("PredictiveEcology/Require@development",
           "PredictiveEcology/SpaDES.project@development"), ask = FALSE)

out <- SpaDES.project::setupProject(
  modules = c(
    "PredictiveEcology/Biomass_borealDataPrep",
    "PredictiveEcology/Biomass_speciesData",
    "PredictiveEcology/Biomass_core"
  ),
  overwrite = TRUE,
  times = list(start = 1, end = 1),
  packages = c("PredictiveEcology/LandR@development (>= 1.2.0.9002)"),
  shapeZipID = "1gycTXyZzgXIUGM6dyAJbmFxq8TOAm-Ik",
  studyArea = reproducible::prepInputs(url = shapeZipID, fun = terra::vect) |>
    terra::project("EPSG:3978"),
  studyAreaLarge = studyArea
)

finalOut <- SpaDES.core::simInitAndSpades2(out)
```

Roughly ninety lines become fifteen, but the point is not brevity for
its own sake -- it is *what each line now does*.

## What changed, step by step

### Packages: from `library()` to a managed library

The hand-rolled version `library()`s a fixed set of packages and assumes
they are already installed at compatible versions. The nimble version
hands a `packages` argument to `setupProject()`, which **installs and
loads** them into a project-specific library.

The important part is *where* this happens. Because `setupProject()`
also knows the `modules`, it can read each module's own package
requirements (declared in their metadata) and reconcile them **together
with the packages you ask for** -- in one resolution step, before
anything runs. So a module that needs a particular version of `LandR`
and your own analysis needs are satisfied at the same time, rather than
discovered one failed `library()` call at a time. Note we only list
`LandR` explicitly here; `SpaDES.core`, `terra` and the rest arrive as
dependencies of the modules and packages requested.

### Paths: from `setPaths()` + manual cache cleanup to defaults

The script sets five paths by hand and then deletes and recreates the
cache to "avoid previous mismatches". `setupProject()` establishes a
sensible project directory layout for you (override it with a `paths`
argument when you need to). And deleting the cache to avoid mismatches
is treating a symptom: `Cache` keys on its inputs, so a correct cache
*doesn't* go stale (see [Introduction to Cache @sec-caching]) -- there is
nothing to clear.

### Getting the modules

`getModule()` becomes the `modules` argument. `setupProject()` downloads
the listed modules from their repositories as part of setup. (How
modules let you *compose* a workflow is the subject of the other
chapters in this section; here we are only moving the download step.)

### Data: from `download.file()` + `unzip()` + `vect()` to `prepInputs()`

This is the heart of the translation. The manual version spends some
forty lines turning a Google Drive id into an R object: build a URL,
create a directory, conditionally download, conditionally unzip, glob
for the `.shp`, assert there is exactly one, read it, reproject it.
Every one of those steps is a place to special-case, to forget, or to
get subtly wrong on a different machine.

`reproducible::prepInputs()` collapses all of it:

```{r prepinputs-spotlight}
studyArea = reproducible::prepInputs(url = shapeZipID, fun = terra::vect) |>
  terra::project("EPSG:3978")
```

`prepInputs()` downloads from the `url`, unzips if needed, locates and
reads the file with `fun`, and -- because it is built on `Cache` -- does
the download and processing **once**, reusing the result on every
subsequent run (see [`prepInputs` @sec-prepInputs]). The decisive shift
is that the study area now **starts from a URL, not a local file path**.
Nothing in the workflow assumes a file already exists on disk, so it
runs the same way on any machine, for anyone, the first time.

Two smaller things worth noticing:

-   `shapeZipID` is passed as an argument to `setupProject()` and then
    *referenced* by `studyArea`. Arguments are evaluated in order, and
    each becomes available to those after it -- which is also why
    `studyAreaLarge = studyArea` works. The whole project description
    stays in one self-contained call.
-   The reprojection (`terra::project(..., "EPSG:3978")`) is piped
    directly onto the loaded object, so "the study area" is *defined* as
    "this remote file, read and reprojected" -- a recipe, not a file.

### Settings and objects

`times` is passed straight through. The `objects` list disappears:
`studyArea` and `studyAreaLarge` are simply named arguments now, and
`setupProject()` routes named objects, `times`, `paths`, `params`,
`modules` and `packages` to the right place for you.

### Running: `simInit()` + `spades()` → `simInitAndSpades2()`

`setupProject()` returns a list of fully-prepared arguments. The `2`
variant of `simInitAndSpades()` accepts that list directly, so the
two-step `simInit()` then `spades()` becomes a single
`simInitAndSpades2(out)`.

## Why "nimble"?

The translated script is *nimble* because it can move:

-   **Portable** -- every input starts from a URL or repository, so
    there are no machine-specific file paths to fix up.
-   **Self-installing** -- packages (yours and the modules') are
    resolved and installed together, up front.
-   **Reproducible and fast on re-run** -- `prepInputs()` and `Cache`
    avoid re-downloading and re-computing unchanged steps.
-   **Declarative** -- the entire project is one call describing *what*
    is needed, not a sequence of imperative bookkeeping that must
    succeed in order.

These same properties are what let the richer projects in the following
chapters chain many modules together without the setup collapsing under
its own weight.

## See also

[Workflows with `setupProject`](workflowSimple2.qmd)

[`reproducible::prepInputs` for Data @sec-prepInputs]

[Introduction to Cache @sec-caching]

[`?SpaDES.project::setupProject`](https://spades-project.predictiveecology.org/reference/setupProject.html)

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path("R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
