---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  eval: false
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# `setupProject` for nimble workflows {#sec-setupProjectNimble}

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

Most SpaDES workflows begin life as an ordinary R script: load some
packages, set some paths, download a few things, read a shapefile,
initialize a simulation, run it. That script *works* -- on the machine
where it was written. Moving it to another computer (a colleague's, a
cluster, the version of your own machine six months from now) is where
the trouble starts: a missing package, a hard-coded path that doesn't
exist, a data file that was never in the repository.

This chapter takes one such script and translates it, step by step, into
the equivalent `SpaDES.project::setupProject()` call. The simulation it
runs is identical. What changes is that the *nimble* version is
reproducible and portable by construction: everything it needs is
described declaratively, packages are reconciled in one place, and every
input **starts from a URL, not a local file path**.

## The starting point: a hand-rolled script

Here is an actual example that was written by a student. It is a 
"works on my machine" setup for a small `LandR`
Biomass run. It is not wrong -- it is just doing a lot of bookkeeping by
hand, and most of that bookkeeping is exactly what breaks elsewhere.

```{r before}
## 1) Packages -------------------------------------------------------
## to be portable, we must INSTALL what is missing ...
if (!requireNamespace("SpaDES.core",    quietly = TRUE)) install.packages("SpaDES.core")
if (!requireNamespace("SpaDES.project", quietly = TRUE)) install.packages("SpaDES.project")
if (!requireNamespace("terra",          quietly = TRUE)) install.packages("terra")
if (!requireNamespace("sf",             quietly = TRUE)) install.packages("sf")

## ... and then still LOAD it (installing does not attach a package)
library(SpaDES.core)
library(SpaDES.project)
library(terra)
library(sf)

## 2) Paths ----------------------------------------------------------
setPaths(
  modulePath  = "~",
  inputPath   = "~",
  outputPath  = "~",
  cachePath   = "~",
  scratchPath = "~"
)

## clean cache to avoid previous mismatches
unlink(getPaths()$cachePath, recursive = TRUE)
dir.create(getPaths()$cachePath, showWarnings = FALSE)

## 3) Download modules ----------------------------------------------
SpaDES.project::getModule(
  modules = c(
    "PredictiveEcology/Biomass_borealDataPrep",
    "PredictiveEcology/Biomass_speciesData",
    "PredictiveEcology/Biomass_core"
  ),
  modulePath = getPaths()$modulePath,
  overwrite  = TRUE
)

## 4) Download study area from Google Drive -------------------------
shapeZipID <- "1gycTXyZzgXIUGM6dyAJbmFxq8TOAm-Ik"
shapeDir   <- file.path(getPaths()$inputPath, "studyArea")
dir.create(shapeDir, recursive = TRUE, showWarnings = FALSE)
shapeZip   <- file.path(shapeDir, "studyArea.zip")

if (!file.exists(shapeZip)) {
  download.file(
    url      = paste0("https://drive.google.com/uc?export=download&id=", shapeZipID),
    destfile = shapeZip,
    mode     = "wb"
  )
}

## unzip ------------------------------------------------------------
if (length(list.files(shapeDir, pattern = "\\.shp$", recursive = TRUE)) == 0) {
  unzip(shapeZip, exdir = shapeDir)
}

## read shapefile ---------------------------------------------------
shpFile <- list.files(shapeDir, pattern = "\\.shp$",
                      recursive = TRUE, full.names = TRUE)
stopifnot(length(shpFile) == 1)

studyArea <- terra::vect(shpFile)
studyArea <- terra::project(studyArea, "EPSG:3978")

## 5) Simulation settings -------------------------------------------
times   <- list(start = 1, end = 1)
modules <- c("Biomass_borealDataPrep", "Biomass_speciesData", "Biomass_core")
objects <- list(studyArea = studyArea, studyAreaLarge = studyArea)

## 6) Initialize and run --------------------------------------------
sim <- simInit(times = times, modules = modules,
               objects = objects, paths = getPaths())
sim <- spades(sim)
```

## The same workflow with `setupProject`

The same run, expressed declaratively:

```{r after}
if (!require("pak")) install.packages("pak")
pak::pak(c("PredictiveEcology/Require@development",
           "PredictiveEcology/SpaDES.project@development"), ask = FALSE)

out <- SpaDES.project::setupProject(
  modules = c(
    "PredictiveEcology/Biomass_borealDataPrep",
    "PredictiveEcology/Biomass_speciesData",
    "PredictiveEcology/Biomass_core"
  ),
  overwrite = TRUE,
  times = list(start = 1, end = 1),
  packages = c("PredictiveEcology/LandR@development (>= 1.2.0.9002)"),
  shapeZipID = "1gycTXyZzgXIUGM6dyAJbmFxq8TOAm-Ik",
  studyArea = reproducible::prepInputs(url = shapeZipID, fun = terra::vect) |>
    terra::project("EPSG:3978"),
  studyAreaLarge = studyArea
)

finalOut <- SpaDES.core::simInitAndSpades2(out)
```

Roughly ninety lines become fifteen, but the point is not brevity for
its own sake -- it is *what each line now does*.

## What changed, step by step

### Packages: installing what's missing, not just loading

A bare `library(SpaDES.core)` is not portable: it fails on any machine
where the package isn't already installed. So for the hand-rolled script
to actually run for *someone else*, it cannot simply `library()` -- it
has to **install what is missing** *and then load it*:

```{r install-guard}
## install if missing ...
if (!requireNamespace("SpaDES.core",    quietly = TRUE)) install.packages("SpaDES.core")
if (!requireNamespace("SpaDES.project", quietly = TRUE)) install.packages("SpaDES.project")
if (!requireNamespace("terra",          quietly = TRUE)) install.packages("terra")
if (!requireNamespace("sf",             quietly = TRUE)) install.packages("sf")
## ... and still load, because installing does not attach
library(SpaDES.core)
library(SpaDES.project)
library(terra)
library(sf)
```

Note that the install and the load are *separate* steps: installing a
package does not attach it. This is why the popular one-liner
`if (!require(x)) install.packages(x)` is subtly broken on a fresh
machine -- `require()` only loads `x` when it is *already* installed, so
on the very run where it installs `x`, it never loads it. Getting even
this small dance right, for every package, is its own little chore.

That aside, the guard-and-load approach has three deeper weaknesses, and
they grow with the project:

1.  **Repetitive** -- two lines per package, by hand, kept in sync
    forever.
2.  **Wrong source** -- `install.packages()` uses whatever repositories
    happen to be configured, so the PredictiveEcology packages (which
    live on the `r-universe` repo) and any GitHub-only or
    version-pinned packages won't be found without extra setup.
3.  **Incomplete** -- it only installs *your* packages. It has no idea
    which packages the **modules** need, or at what versions, so those
    failures surface one at a time, deep into a run.

The nimble version replaces all of this with the `packages` argument to
`setupProject()`, which **installs and loads** the requested packages
into a project-specific library. It still needs a tiny bootstrap -- the
same `if (!require(...))` idiom -- but only to install the *one* tool
that does the installing:

```{r pak-bootstrap}
if (!require("pak")) install.packages("pak")
pak::pak(c("PredictiveEcology/Require@development",
           "PredictiveEcology/SpaDES.project@development"), ask = FALSE)
```

After that, everything is declarative. And the important part is *where*
the resolution happens: because `setupProject()` also knows the
`modules`, it reads each module's own package requirements (declared in
their metadata) and reconciles them **together with the packages you ask
for** -- in one step, before anything runs. So a module that needs a
particular version of `LandR` (note the `@development (>= 1.2.0.9002)`
pin) and your own analysis needs are satisfied at once, rather than
discovered one failed `library()` call at a time. We only list `LandR`
explicitly; `SpaDES.core`, `terra` and the rest arrive as dependencies
of the requested modules and packages.

### Paths: from `setPaths()` + manual cache cleanup to defaults

The script sets five paths by hand and then deletes and recreates the
cache to "avoid previous mismatches". `setupProject()` establishes a
sensible project directory layout for you (override it with a `paths`
argument when you need to). And deleting the cache to avoid mismatches
is treating a symptom: `Cache` keys on its inputs, so a correct cache
*doesn't* go stale (see [Introduction to Cache @sec-caching]) -- there is
nothing to clear.

### Getting the modules

`getModule()` becomes the `modules` argument. `setupProject()` downloads
the listed modules from their repositories as part of setup. (How
modules let you *compose* a workflow is the subject of the other
chapters in this section; here we are only moving the download step.)

### Data: from `download.file()` + `unzip()` + `vect()` to `prepInputs()`

This is the heart of the translation. The manual version spends some
forty lines turning a Google Drive id into an R object: build a URL,
create a directory, conditionally download, conditionally unzip, glob
for the `.shp`, assert there is exactly one, read it, reproject it.
Every one of those steps is a place to special-case, to forget, or to
get subtly wrong on a different machine.

`reproducible::prepInputs()` collapses all of it:

```{r prepinputs-spotlight}
studyArea = reproducible::prepInputs(url = shapeZipID, fun = terra::vect) |>
  terra::project("EPSG:3978")
```

`prepInputs()` downloads from the `url`, unzips if needed, locates and
reads the file with `fun`, and -- because it is built on `Cache` -- does
the download and processing **once**, reusing the result on every
subsequent run (see [`prepInputs` @sec-prepInputs]). The decisive shift
is that the study area now **starts from a URL, not a local file path**.
Nothing in the workflow assumes a file already exists on disk, so it
runs the same way on any machine, for anyone, the first time.

Two smaller things worth noticing:

-   `shapeZipID` is passed as an argument to `setupProject()` and then
    *referenced* by `studyArea`. Arguments are evaluated in order, and
    each becomes available to those after it -- which is also why
    `studyAreaLarge = studyArea` works. The whole project description
    stays in one self-contained call.
-   The reprojection (`terra::project(..., "EPSG:3978")`) is piped
    directly onto the loaded object, so "the study area" is *defined* as
    "this remote file, read and reprojected" -- a recipe, not a file.

### Settings and objects

`times` is passed straight through. The `objects` list disappears:
`studyArea` and `studyAreaLarge` are simply named arguments now, and
`setupProject()` routes named objects, `times`, `paths`, `params`,
`modules` and `packages` to the right place for you.

### Running: `simInit()` + `spades()` → `simInitAndSpades2()`

`setupProject()` returns a list of fully-prepared arguments. The `2`
variant of `simInitAndSpades()` accepts that list directly, so the
two-step `simInit()` then `spades()` becomes a single
`simInitAndSpades2(out)`.

### Why not just put everything in the `.GlobalEnv`?

The hand-rolled script creates `studyArea`, `times`, `objects` and the
rest as ordinary variables in your global workspace, the `.GlobalEnv`.
For a single, short script that is harmless. But the `.GlobalEnv` is
special in R: it sits at the top of the search path, so **everything can
see it** -- any function, any module, anywhere, can reach an object that
lives there.

That sounds convenient, and that is exactly the problem. When a module
goes looking for an object and *doesn't* find it where it should -- in
the `simList` -- R does not stop with an error. It keeps searching
outward and, if a variable of the same name happens to be sitting in the
`.GlobalEnv`, it quietly uses that one instead and carries on. The run
appears to "work", but for the wrong reason: a genuinely missing or
misnamed input has been silently papered over by a leftover global.

This has bitten us so many times that we treat it as a rule of thumb:
the `.GlobalEnv` is fine for small, simple projects, but as soon as a
project grows past a single script it becomes a common source of
confusing, hard-to-reproduce bugs.

`setupProject()` sidesteps the trap by design. The objects you pass
(`studyArea`, `studyAreaLarge`, ...) are routed into the `simList`
rather than scattered across the global workspace, so a module that asks
for an input it wasn't given fails **loudly**, where you can see and fix
it, instead of silently borrowing whatever the `.GlobalEnv` happened to
contain.

## Why "nimble"?

The translated script is *nimble* because it can move:

-   **Portable** -- every input starts from a URL or repository, so
    there are no machine-specific file paths to fix up.
-   **Self-installing** -- packages (yours and the modules') are
    resolved and installed together, up front.
-   **Reproducible and fast on re-run** -- `prepInputs()` and `Cache`
    avoid re-downloading and re-computing unchanged steps.
-   **Declarative** -- the entire project is one call describing *what*
    is needed, not a sequence of imperative bookkeeping that must
    succeed in order.

These same properties are what let the richer projects in the following
chapters chain many modules together without the setup collapsing under
its own weight.

## See also

[Workflows with `setupProject`](workflowSimple2.qmd)

[`reproducible::prepInputs` for Data @sec-prepInputs]

[Introduction to Cache @sec-caching]

[`?SpaDES.project::setupProject`](https://spades-project.predictiveecology.org/reference/setupProject.html)

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path("R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
