---
author: "Eliot McIntire"
date: last-modified
editor_options:
  chunk_output_type: console
execute:
  tidy: false
  warning: false
  message: false
  error: false
  cache: false
---

# Workflows with `setupProject`

::: {.callout-tip appearance="simple"}
See [Barebones R script] for the code shown in this chapter
:::

```{r setup}
#| include: false
#| eval: true
#| purl: false

# truncate text output
hook_output <- knitr::knit_hooks$get("output")

knitr::knit_hooks$set(output = function(y, options) {
  if (!is.null(n <- options$outlines)) {
    y <- xfun::split_lines(y)
    if (length(y) > n) {
      # truncate the output
      y <- c(head(y, n), "....\n")
    }
    y <- paste(y, collapse = "\n")
  }
  hook_output(y, options)
})

repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
if (!requireNamespace("SpaDES.project", quietly = TRUE))
  install.packages(c("Require", "SpaDES.project"), repos = repos)
```

Once we want to run a project that is more sophisticated, we can start
using `SpaDES.project`.
[`SpaDES.project`](https://spades-project.predictiveecology.org/) is a
package that deals with a number of challenges that crop up creating and
using workflows. These become particularly acute when working with
multiple modules created by potentially many people. [See here to see a
longer list of project
challenges](https://spades-project.predictiveecology.org/#project-challenges).

We will demonstrate its use with an example created by [Tati
Micheletti](https://github.com/tati-micheletti/Micheletti_SpaDES/blob/main/demo.R)
where a workflow with several modules is setup and run.

## `SpaDES.project::setupProject`

https://spades-project.predictiveecology.org/articles/i-getting-started.html

## Using `setupProject`

`setupProject` does a series of very specific things, in a particular
order, so that the chances of package conflicts and user errors are
minimized. It downloads the listed modules directly from their GitHub
repositories (e.g.,
[tati-micheletti/speciesAbundance](https://github.com/tati-micheletti/speciesAbundance))
and installs the listed packages into a project-specific library.

```{r setupProject}
# If SpaDES.project is not installed, see:
# https://predictiveecology.org/training/_book/PackagesForBook.html
repos <- c("https://predictiveecology.r-universe.dev", getOption("repos"))
library(SpaDES.project)
Setup <- setupProject(
  # TODO: simplify once dev branches are merged and released to CRAN/r-universe
  packages = c("PredictiveEcology/Require@usePak",
               "PredictiveEcology/reproducible@recovery2",
               "PredictiveEcology/SpaDES.project@working/combined-prs",
               "PredictiveEcology/SpaDES.core@fixRCMDcheckWarnings",
               "terra"),
  modules = c("tati-micheletti/speciesAbundance@main",
              "tati-micheletti/temperature@main",
              "tati-micheletti/speciesAbundTempLM@main"),
  options = list(repos = repos),
  paths = list(projectPath = "~/SpaDES_book/workflowSimple2"),
  times = list(start = 2013, end = 2030),
  updateRprofile = FALSE
)
```

The output of `setupProject` is a list of argument values that can be
passed to `simInitAndSpaDES2` (with the "2" it accepts a list of
arguments, instead of the arguments themselves).

```{r runSimInit}
results <- SpaDES.core::simInitAndSpades2(Setup)
```

## Emergent workflow

We can evaluate the sequence of events using `completed`

```{r evaluteResults}
SpaDES.core::completed(results)
```

We can also see how modules and objects interact using the functions
objectDiagram() and moduleDiagram().

Modules’ diagram showing the interactions among modules:

```{r moduleDiagram}
SpaDES.core::moduleDiagram(results)
```

We can also access any objects and use them. For example, we can look at
the forecasted abundances and the difference raster is also possible.
This is done by calling the object name from the results list.

```{r objectDiagram}
terra::plot(terra::rast(results$forecasts))
```

## Examples

In the remaining chapters of this section, we will explore several
examples, including learning of several arguments to the function
`setupProject`. But we will start very simple first.

## See also

[All Functions in
SpaDES.core](https://spades-core.predictiveecology.org/reference/index.html)

[?SpaDES.core::completed](https://spades-core.predictiveecology.org/reference/simList-accessors-events.html)

## Barebones R script

```{r}
#| eval: true
#| echo: false
#| purl: false
#| cache: false

fileName <- file.path("R/Chapter_scripts", sub("rmarkdown", "R", knitr::current_input()))
```

```{r file = fileName}
#| eval: false
#| echo: true
#| purl: false
#| cache: false

```
